---
title: "Chapter 3 Exercises"
output: html_notebook
---
```{r}
#if caret and mlbench aren't installed, install and squelch messages
if (!require("caret",quietly=TRUE)) install.packages("caret")
if (!require("mlbench",quietly=TRUE)) install.packages("mlbench")
#attach these packages and squelch message
suppressPackageStartupMessages(library(mlbench))
#call the data from the mlbench library
data(Glass)
#display  the structure of the dataset
str(Glass)
```
## Part 3.1
### A) *Using visualizations, explore the predictor variables to understand their distrubitons as well as the relationships betewen predictors.*

The code chunk below subsets the columns containing numeric values and places them, by column(elements that make glass), into a scatter plot matrix in order to get a rough estimation of linearity between predicator sets.
```{r}
#subset numeric values in all rows, columns 1-9
num<-Glass[,1:9]
#create a table of scatter plots
plot(num,main="Scatterplot Matrix of Numeric Predictors") 
```
#### Analyis of Scatterplot Matrix
Based on the figure output by the plot function above, it appears that there is a strong positive correlation between the variable pair of RI and Ca. The pairset RI and Si seem to have a weak or moderate negative correlation based on their respective graphs. The rest of the variable sets appaear to have no correlation between them.



```{r}
par(mfrow = c(3,3)) #set parameter for graphics to 3x3 figure
for (x in 1:ncol(num)) {
  hist(num[ ,x], xlab = names(num[x]), main = paste(names(num[x]), "Histogram"))
} #iterate through columns in num creating histograms of data and naming them as the column name + Histogram
```












